---
title: "Multivariable Logistic Regression Analysis of Diabetes and Modifiable Risk Factors for Heart Disease: 2022 BRFSS Data"
author: "Claire Xu"
format:
  pdf:
    documentclass: article
    geometry:
      - margin= 1 in
    fontsize: 12pt
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references_heart_disease.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gtsummary)
library(broom)
library(pROC)
library(corrplot)
library(ResourceSelection)
library(dplyr)
library(stringr)
library(knitr)
library(kableExtra) 
library(flextable)

# Sets the default ggplot2 theme to a clean black-and-white theme with base font size 10 for all subsequent plots
theme_set(theme_bw(base_size = 10))  

# Load the data 
heart <- read_csv("/Users/clairexu/Desktop/Heart Disease Project/csv files/2022/heart_2022_no_nans.csv")

# Recoding/factoring: 

# Recode heart attack status as a binary outcome
heart <- heart |>
  mutate(HeartDisease = if_else(HadHeartAttack == "Yes", 1, 0))

# Recode HadDiabetes into binary yes/no
heart <- heart|>
  mutate(
    HadDiabetes_binary = case_when(
      HadDiabetes == "Yes" ~ "Yes",
      HadDiabetes == "No" ~ "No",
      HadDiabetes == "No, pre-diabetes or borderline diabetes" ~ "No",
      HadDiabetes == "Yes, but only during pregnancy (female)" ~ NA_character_
    ),
    HadDiabetes_binary = factor(HadDiabetes_binary, levels = c("No", "Yes"))
  )

# Recode all remaining covariates as factors
heart<- heart |>
  mutate(
    # factor version of outcome for table display
    HeartDisease_f = factor(HeartDisease, levels = c(0, 1),
                            labels = c("No", "Yes")),
    # Self-reported health (Excellent = reference)
    GeneralHealth = factor(
      GeneralHealth,
      levels = c("Excellent", "Very good",
                 "Good", "Fair", "Poor")
    ),
    
    # Age groups (youngest group = reference)
    AgeCategory = factor(
      AgeCategory,
      levels = c("Age 18 to 24", "Age 25 to 29",
                 "Age 30 to 34", "Age 35 to 39",
                 "Age 40 to 44", "Age 45 to 49",
                 "Age 50 to 54", "Age 55 to 59",
                 "Age 60 to 64", "Age 65 to 69",
                 "Age 70 to 74", "Age 75 to 79",
                 "Age 80 or older")
    ),
    
    # Smoking history (never smoked = reference)
    SmokerStatus = factor(
      SmokerStatus,
      levels = c("Never smoked",
                 "Former smoker",
                 "Current smoker - now smokes every day",
                 "Current smoker - now smokes some days")
    ),
    
    # Binary covariates
    PhysicalActivities = factor(PhysicalActivities, levels = c("No", "Yes")),
    Sex = factor(Sex, levels = c("Female", "Male")),
    
    #Race/ethnicity (White only, Non-Hispanic" = reference)
    RaceEthnicityCategory = factor(
      RaceEthnicityCategory,
      levels = c("White only, Non-Hispanic",
                 "Black only, Non-Hispanic",
                 "Hispanic",
                 "Other race only, Non-Hispanic",
                 "Multiracial, Non-Hispanic")
    )
  )

# Create BMI categories using standard WHO cutoffs
heart <- heart |>
  mutate(
    BMICategory = case_when(
      BMI < 18.5 ~ "Underweight",
      BMI < 25 ~ "Normal",
      BMI < 30 ~ "Overweight",
      BMI >= 30 ~ "Obese"
    ),
    # Reference level = Normal
    BMICategory = factor(
      BMICategory,
      levels = c("Normal", "Underweight", "Overweight", "Obese")
    )
  )

# Filter out NA values for HadDiabetes before fitting
heart_complete <- heart |> filter(!is.na(HadDiabetes_binary))

# Setting up all the models:
# primary model with complete case
model_rq1 <- glm(
  HeartDisease ~ HadDiabetes_binary + AgeCategory + Sex + BMI +
    SmokerStatus,
  data = heart_complete,
  family = binomial
)

# Model specification: 

# Full model with all covariates including optional ones
model_full <- glm(
  HeartDisease ~ HadDiabetes_binary + AgeCategory + Sex + BMI +
    SmokerStatus + PhysicalActivities + GeneralHealth +
    RaceEthnicityCategory + SleepHours + AlcoholDrinkers +
    HadDepressiveDisorder + DifficultyWalking,
  data = heart,
  family = binomial
)

# Female only model
model_female <- glm(HeartDisease ~ HadDiabetes_binary + AgeCategory +
                      BMI + SmokerStatus,
                    data = subset(heart_complete, Sex == "Female"),
                    family = binomial)

# Male only model
model_male <- glm(HeartDisease ~ HadDiabetes_binary + AgeCategory
                    + BMI + SmokerStatus, 
                    data = subset(heart_complete, Sex == "Male"),
                    family = binomial)

# Predicted probabilities form primary and full model
pred_rq1 <- predict(model_rq1, type = "response")
pred_full <- predict(model_full, type = "response")

# ROC curves
roc_rq1 <- roc(heart_complete$HeartDisease, pred_rq1)
roc_full <- roc(heart_complete$HeartDisease, pred_full)

# Find the optimal cutoff using Youden's Index
youden <- coords(
  roc_rq1,
  "best",                         
  best.method = "youden",         
  ret = c("threshold", "sensitivity", "specificity")          
)

# Sensitivity analysis:

# Recode HadDiabetes in a full dataset keeping gestational as separate level
heart_sensitivity <- heart |>
  mutate(
    HadDiabetes_3level = case_when(
      HadDiabetes == "Yes" ~ "Yes",
      HadDiabetes == "No" ~ "No",
      HadDiabetes == "No, pre-diabetes or borderline diabetes" ~ "No",
      HadDiabetes == "Yes, but only during pregnancy (female)" ~ "Gestational"
    ),
    HadDiabetes_3level = factor(
      HadDiabetes_3level,
      levels = c("No", "Yes", "Gestational")
    )
  )

# Refit primary model with 3-level diabetes variable
model_sensitivity <- glm(
  HeartDisease ~ HadDiabetes_3level + AgeCategory +
    Sex + BMI + SmokerStatus,
  data = heart_sensitivity,
  family = binomial
)

# Bootstrap AUC: load saved result instead of rerunning
boot_auc <- readRDS("boot_auc.rds")


# Setting up dataframes:

# Add fitted values and residuals to a data frame
# Run diagnostics for the primary model
model_rq1_diag <- data.frame(
  fitted = fitted(model_rq1),
  deviance_resid = residuals(model_rq1, type = "deviance"),
  pearson_resid = residuals(model_rq1, type = "pearson")
)

# Create data frame for forest plot
coef_table <- summary(model_rq1)$coefficients
OR <- exp(coef(model_rq1))
lower <- exp(coef(model_rq1) - 1.96 * coef_table[, "Std. Error"])
upper <- exp(coef(model_rq1) + 1.96 * coef_table[, "Std. Error"])

forest_data <- data.frame(
  variable = names(OR),
  OR = OR,
  lower = lower,
  upper = upper
)

# Compute influence measures in dataframe
influence_measures <- data.frame(
  cooks = cooks.distance(model_rq1),
  hat = hatvalues(model_rq1)
)

```

# Abstract

Heart disease remains one of the leading causes of morbidity and
mortality in the United States, making the identification of major risk
factors an important public health objective. This study used data from
the 2022 Behavioral Risk Factor Surveillance System (BRFSS) to examine
the association between diabetes and heart disease while accounting for
demographic and behavioral risk factors. The full dataset included
246,022 respondents. The primary analysis used a complete-case sample of
244,039 participants after excluding 1,983 respondents with missing
diabetes values created during the recoding of gestational diabetes.

Multivariable logistic regression was used to estimate crude and
adjusted odds ratios (ORs), along with 95% confidence intervals (CIs).
Additional analyses examined physical activity, smoking, self-reported
general health, and whether the association between smoking and heart
disease differed by age. Model performance was evaluated using the
receiver operating characteristic (ROC) curve, with the area under the
curve (AUC) used to summarize the model's ability to correctly
distinguish between participants with and without heart disease.
Internal validation was performed using bootstrap resampling to evaluate
how well the model would generalize to similar populations. Diabetes was
independently associated with more than twice the odds of heart disease
(adjusted OR = 2.339, 95% CI: 2.247–2.435). A sensitivity analysis that
included respondents with gestational diabetes produced nearly identical
results (adjusted OR = 2.340, 95% CI: 2.248–2.436), indicating that the
primary findings were robust. Older age, male sex, current smoking,
physical inactivity, and poorer self-reported general health were also
significantly associated with higher odds of heart disease. The primary
model showed good predictive performance (AUC = 0.786), and bootstrap
validation showed minimal optimism, suggesting little evidence of
overfitting. Overall, these findings reinforce diabetes as a strong
independent predictor of heart disease and emphasize the importance of
prevention, early detection, and management of modifiable cardiovascular
risk factors.

# Introduction

Cardiovascular disease (CVD), especially coronary heart disease (CHD),
remains the leading cause of death in the United States, representing a
major public health burden due to its high prevalence, mortality, and
associated healthcare costs [@shah2025]. Established risk factors for
heart disease include diabetes, smoking, obesity, physical inactivity,
and other behavioral and demographic characteristics [@leal2024]. While
these risk factors are well documented, their relative contributions can
vary across populations depending on clinical, lifestyle, and
socioeconomic factors. Prior BRFSS-based research found that adults with
diabetes alone (without co-occurring metabolic syndrome) had higher odds
of heart attack compared to those with neither diabetes nor metabolic
syndrome, even after adjusting for demographic and behavioral
confounders [@yang2019].

Large surveillance systems such as the Behavioral Risk Factor
Surveillance System (BRFSS) are widely used to study cardiovascular risk
factors in population-based samples [@yang2019; @leal2024; @shah2025;
@yathindra2025]. The BRFSS dataset has a large sample size, which helps
detect small effects and compare different subgroups. However, it also
has limitations, including reliance on self-reported measures for both
exposures and outcomes, as well as its cross-sectional design, which
prevents determination of temporal ordering. Even with these
limitations, large population health datasets such as BRFSS can support
prevention efforts and inform public health interventions when
interpreted alongside existing literature.

This study uses 2022 BRFSS data to examine whether diabetes is
associated with higher odds of heart disease after adjusting for age,
sex, BMI, and smoking status. It is hypothesized that adults with a
diabetes diagnosis will have higher odds of heart disease than
non-diabetic adults after adjustment, consistent with prior BRFSS
research linking diabetes to elevated cardiovascular risk [@yang2019].

Three secondary questions were also examined. First, whether physical
inactivity independently predicts heart disease after adjusting for BMI
and age, given evidence that behavioral risk factors including physical
inactivity are associated with elevated cardiovascular odds [@shah2025].
Second, whether the association between smoking and heart disease
differs by age group, with the expectation that this association is
stronger among older adults due to cumulative exposure, building on
findings that smoking-related cardiovascular risk varies meaningfully
across age groups [@yathindra2025]. Third, whether general health status
is associated with heart disease after adjusting for behavioral and
clinical factors, with the expectation that odds increase progressively
from excellent to poor self-reported health.

The objective of this study is to estimate these associations using
multivariable logistic regression on a large population-based sample,
evaluate model performance, and interpret the results in the context of
existing literature on cardiovascular risk.

# Data and Methods

## Data Source

This study uses the *Indicators of Heart Disease* dataset (2022),
derived from the Behavioral Risk Factor Surveillance System (BRFSS),
administered by the Centers for Disease Control and Prevention
[@cdc_brfss]. The data were accessed through Kaggle [@kaggle_dataset]
and contain a cleaned subset of 41 variables and 246,022 observations
selected from the original BRFSS survey, which includes over 300
variables capturing demographic characteristics, health behaviors, and
chronic conditions related to cardiovascular health. The dataset
contains no missing values, so all 246,022 observations were initially
available for analysis.

## Study Population

The study population consists of U.S. adults aged 18 and older who
participated in the 2022 BRFSS telephone survey. For the primary and
subgroup analysis, 1,983 respondents who reported diabetes only during
pregnancy were excluded, as gestational diabetes is a distinct clinical
condition from diabetes diagnoses of interest in this study and was
similarly excluded in prior BRFSS-based research [@yang2019].
Respondents reporting pre-diabetes or borderline diabetes were retained
and classified as non-diabetic, consistent with their absence of a
confirmed diabetes diagnosis. After these exclusions, the complete-case
sample for the primary analysis consisted of 244,039 respondents,
including only respondents with complete data for all variables used in
the analysis. A sensitivity analysis was conducted using the full sample
of 246,022 respondents, with gestational diabetes coded as a separate
category, to assess whether this exclusion meaningfully affected the
primary results.

## Outcome and Exposure Variables

The outcome variable was `HeartDisease`, a binary variable indicating
whether a respondent had ever been diagnosed with a heart attack. This
variable was derived from the survey item `HadHeartAttack` and recoded
as Yes = 1 and No = 0.

The primary exposure variable was diabetes status, derived from the
original four-level `HadDiabetes` variable. Diabetes status was recoded
into a binary variable (Yes = 1, No = 0) for analysis. Respondents
reporting gestational diabetes were excluded because gestational
diabetes occurs during pregnancy and is clinically distinct from type 1
and type 2 diabetes. Respondents reporting pre-diabetes or borderline
diabetes were classified as non-diabetic because the primary analysis
focused on diagnosed diabetes. This binary coding simplified
interpretation of the association between diabetes and heart disease.

## Covariates

Covariates were selected based on established cardiovascular risk
factors identified in the literature and the secondary research
questions of interest. These included age category, sex, body mass index
(BMI), smoking status, self-reported general health, physical activity,
and race/ethnicity.

Age category was treated as an ordered factor with 13 levels (18–24
years through 80 years and older), using the youngest age group as the
reference category. Sex was coded as a binary factor, with female as the
reference category. Body mass index (BMI) was analyzed as a continuous
variable and was also categorized into four groups (Underweight, Normal
weight, Overweight, and Obese) using standard World Health Organization
cutoffs for descriptive analyses. Smoking status was coded as a
four-level factor, with never smoked as the reference category, followed
by former smoker, current smoker every day, and current smoker some
days. Self-reported general health was treated as an ordered five-level
factor, with Excellent as the reference category and Poor representing
the lowest level of health. Physical activity was coded as a binary
factor, with no physical activity as the reference category. Race and
ethnicity were coded as a five-level factor, with White, non-Hispanic as
the reference category.

Additional variables considered in supplementary analyses included sleep
duration, alcohol use, history of depressive disorder, and difficulty
walking.

## Statistical Analysis

Descriptive statistics were used to summarize the study population
overall and stratified by heart disease status. Continuous variables
were described using means and standard deviations, and categorical
variables were summarized using frequencies and percentages. Univariable
associations between heart disease and each covariate were assessed
using chi-square tests for categorical variables and t-tests for
continuous variables.

Multivariable logistic regression was used to estimate adjusted odds
ratios (ORs) with 95% confidence intervals (CIs) for the association
between diabetes and heart disease, adjusting for age, sex, BMI, and
smoking status. The primary logistic regression model was specified as:
$$
\begin{aligned}
\text{logit}(P(\text{HeartDisease} = 1)) = \beta_0 
&+ \beta_1(\text{HadDiabetes}) + \beta_2(\text{AgeCategory})+ \beta_3(\text{Sex}) \\
&+ \beta_4(\text{BMI}) + \beta_5(\text{SmokerStatus})
\end{aligned}
$$

where $\beta_1$ represents the difference in log odds of having heart
disease between individuals with diabetes and those without diabetes,
assuming the other predictors remain constant.

Additional multivariable models were fit to address the secondary
research questions, including a model examining physical activity
adjusted for BMI and age, a model examining general health status
adjusted for behavioral and clinical covariates, and a model including a
multiplicative interaction term between smoking status and age category
to formally test effect modification.

Multicollinearity was assessed informally through pairwise correlations
among covariates. Model fit was evaluated using the Hosmer-Lemeshow
goodness-of-fit test, deviance and Pearson residual plots, and influence
diagnostics based on Cook's distance and leverage (hat values). Model
discrimination was evaluated using receiver operating characteristic
(ROC) curve analysis and the area under the curve (AUC). The optimal
classification threshold was identified using Youden's index to account
for class imbalance in the outcome variable. Internal validation was
conducted using bootstrap resampling with 200 iterations to estimate
optimism in model performance. Subgroup analyses stratified by sex were
conducted to assess whether the association between diabetes and heart
disease differed between male and female respondents. A sensitivity
analysis was conducted to evaluate the impact of excluding gestational
diabetes cases on the primary effect estimate.

All analyses were conducted in R (version 4.4.1), using the `tidyverse`
package for data wrangling and preparation, `gtsummary` for descriptive
tables, `glm()` for logistic regression, `broom` for model output
extraction, `pROC` for ROC and AUC estimation, `corrplot` for
correlation visualization, and `ResourceSelection` for the
Hosmer-Lemeshow test. The full analysis was written in Quarto to ensure
complete reproducibility, with all code provided in the appendix.

# Results

## Descriptive Statistics

The complete-case sample for the primary analysis included 244.039
respondents after excluding 1,983 individuals who reported gestational
diabetes. Among these respondents, 13,435 (5.46%) reported a history of
heart attack, indicating substantial class imbalance in the outcome
variable. This type of class imbalance is common in population-based
cardiovascular studies and was taken into account when evaluating model
performance.

@tbl-descriptive presents descriptive statistics for the study
population overall and stratified by heart disease status. Respondents
with heart disease were older on average and more frequently reported
diabetes, current smoking, physical inactivity, and poorer self-rated
general health compared to respondents without heart disease. Mean BMI
was slightly higher among respondents with heart disease (29.5 vs.
28.6), while sleep duration did not differ meaningfully between groups
(7.04 versus 7.02 hours). These descriptive findings indicate clear
differences in cardiovascular risk profiles between respondents with and
without heart disease, supporting the need for multivariable adjustment
in subsequent analyses.

```{r}
#| label: tbl-descriptive
#| tbl-cap: "Descriptive Characteristics of Study Population, Overall and Stratified by Heart Disease Status"
#| echo: false
#| latex-options: hold_position

heart_complete |>
  select(HeartDisease_f, HadDiabetes_binary, AgeCategory, Sex, BMI,
         BMICategory, SmokerStatus, PhysicalActivities, GeneralHealth,
         RaceEthnicityCategory, SleepHours, AlcoholDrinkers,
         HadDepressiveDisorder, DifficultyWalking) |>

  # Create the summary table grouped by heart disease status
  tbl_summary(
    # Group the summary by heart disease status
    by = HeartDisease_f,

    # Display more descriptive variable labels
    label = list(
      HadDiabetes_binary ~ "Diabetes",
      AgeCategory ~ "Age Category",
      BMICategory ~ "BMI Category",
      SmokerStatus ~ "Smoking Status",
      PhysicalActivities ~ "Physical Activity",
      GeneralHealth ~ "General Health",
      RaceEthnicityCategory ~ "Race/Ethnicity",
      SleepHours ~ "Sleep Hours",
      AlcoholDrinkers ~ "Alcohol Use",
      HadDepressiveDisorder ~ "Depressive Disorder",
      DifficultyWalking ~ "Difficulty Walking"
    ),

    # Display mean and sd for continuous variables
    # and count and percentage for categorical variables
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),

    # Round continuous variables to one decimal places
    digits = all_continuous() ~ 1,

    # Do not include missing values in the table
    missing = "no"
  ) |>

  # Add a column summarizing all participants for each group
  add_overall() |>
  # Add p-values comparing heart disease groups
  add_p() |>
  # Bold variable labels
  bold_labels() |>
  # Bold statistically significant p-values (p < 0.05)
  bold_p(t = 0.05) |>
  as_flex_table() |>
  flextable::font(fontname = "Times New Roman", part = "all") |>
  # Fix the table layout so column widths remain consistent
  flextable::set_table_properties(layout = "fixed") |>
  # Set the width of each column (in inches) to help the table fit on the page
  flextable::width(width = c(2.3, 1.1, 1.1, 1.1, 0.9), unit = "in")

```

## Univariable Associations

Initial univariable analyses showed that all covariates were
significantly associated with heart disease status. The largest
chi-square statistics were observed for general health ($\chi^2$ =
9,893.9), age category ($\chi^2$ = 7,925.1), and difficulty walking
($\chi^2$ = 6,286.6), indicating strong associations with heart disease.
Diabetes was also strongly associated with heart disease ($\chi^2$ =
5,193.4, p \< 0.001), supporting the primary hypothesis. Smoking status
($\chi^2$ = 2,241.3), physical activity ($\chi^2$ = 1,701.6), and
alcohol use ($\chi^2$ = 1,353.2) were also significantly associated with
heart disease. Although race/ethnicity ($\chi^2$ = 195.92) and
depressive disorder ($\chi^2$ = 138.0) had smaller chi-square
statistics, both remained statistically significant.

Univariable logistic regression produced crude odds ratios for each
covariate before adjustment for potential confounders
(@tbl-univariable). Diabetes was associated with higher odds of heart
disease (Crude OR = 3.69, 95% CI: 3.55–3.83). A strong age-related
pattern was observed, with the odds of heart disease increasing from
1.11 among respondents aged 25–29 years to 41.21 among those aged 80
years and older, compared with the 18–24-year reference group.
Self-reported general health showed a similar trend, with crude odds
ratios increasing from 2.01 for respondents reporting very good health
to 18.83 for those reporting poor health, compared to those in excellent
health. Respondents who reported engaging in physical activity had lower
odds of heart disease (OR = 0.47), indicating that physically active
respondents were less likely to have heart disease. The univariable
results demonstrate consistent and strong associations between heart
disease and key demographic, behavioral, and clinical risk factors,
supporting their inclusion in the multivariable models.

```{r}
#| label: tbl-univariable
#| echo: false
#| latex-options: hold_position

# Covariates for univariate logistic regression
covariates <- c(
  "HadDiabetes_binary", "AgeCategory", "Sex", "BMI",
  "BMICategory", "SmokerStatus", "PhysicalActivities",
  "GeneralHealth", "RaceEthnicityCategory"
)

# Fit one logistic regression model for each covariate using lapply()
# var is each covariate being looped through one at a time
crude_models <- lapply(covariates, function(var) {
  # Create model formula: HeartDisease ~ covariate
  formula <- as.formula(paste("HeartDisease ~", var))
  
  # glm() to fit logistic regression model
  model <- glm(
    formula,
    data = heart,
    family = binomial
  )
  
  # Extract crude odds ratios, confidence intervals, and p-values
  tidy(model, exponentiate = TRUE, conf.int = TRUE) |>
    # Remove the intercept row
    filter(term != "(Intercept)") |>
    # Add the covariate name to the results
    # This creates a new column called "covariate"
    mutate(covariate = var)
})

# Combine all model results into one data frame using bind_rows()
crude_results <- bind_rows(crude_models)

# Results table
crude_results |>
  # Strip out messy covariate name prefixes from the levels to fix overlapping
  mutate(term = stringr::str_remove(term, covariate)) |>
  # Rename RaceEthnicityCategory to Race/Ethnicity
  mutate(covariate = ifelse(covariate == "RaceEthnicityCategory", "Race/Ethnicity", covariate)) |>
  # Clean up formatting of column values (removes underscores)
  mutate(covariate = stringr::str_replace_all(covariate, "_", " ")) |>
  # Select relevant columns
  select(covariate, term, estimate, conf.low, conf.high, p.value) |>
  # Round values
  mutate(
    estimate = round(estimate, 2),
    conf.low = round(conf.low, 2),
    conf.high = round(conf.high, 2),
    p.value = ifelse(p.value <0.001, "<0.001", round(p.value, 3))
  ) |>
  # Display as a clean formatted table
  knitr::kable(
    col.names = c("Covariate", "Term", "Crude OR",
                  "95% CI Lower", "95% CI Upper", "P-value"),
    caption = "Crude Odds Ratios from Univariable Logistic Regression"
  )
```

## Primary Model

The primary multivariable logistic regression model examined whether
diabetes was associated with higher odds of heart disease after
adjusting for age, sex, BMI, and smoking status. Adults with diabetes
had significantly higher odds of heart disease compared to non-diabetic
adults (adjusted OR = 2.34, 95% CI: 2.25–2.44, p \< 0.001),supporting
the primary hypothesis. This adjusted estimate was meaningfully lower
than the crude OR of 3.69 observed in the univariable analysis,
indicating that age, sex, BMI, and smoking status accounted for a
portion of the unadjusted association between diabetes and heart
disease.

Age remained the strongest predictor of heart disease after adjustment,
with odds increasing steadily across age categories and reaching the
highest level among respondents aged 80 years and older compared with
the 18–24 reference group (OR = 33.40). Male sex was associated with
higher odds of heart disease relative to female sex (OR = 2.00). Current
smoking was also strongly associated with heart disease, with higher
odds observed among daily smokers (OR = 2.72) and some-days smokers (OR
= 2.49), while former smokers also showed elevated odds (OR = 1.62)
compared with never smokers. BMI showed a modest positive association
with heart disease, with slightly increased odds per unit increase (OR =
1.02). The primary model identified age, sex, smoking status, BMI, and
diabetes as independent predictors of heart disease, with a strong and
consistent age gradient observed across all models.

@fig-forest displays adjusted odds ratios and 95% confidence intervals
from the primary model. After adjustment, all covariates remained
significantly associated with heart disease, as their confidence
intervals for diabetes, age, sex, and current smoking status did not
include the null value of 1.

```{r}
#| label: fig-forest
#| fig-cap: "Adjusted Odds Ratios from Primary Multivariable Model"
#| echo: false

# Create a forest plot
library(dplyr)

# Extract coefficient table from primary model
coef_table <- summary(model_rq1)$coefficients

# Odds ratios
OR <- exp(coef(model_rq1))

# 95% confidence intervals for odds ratios
lower <- exp(coef(model_rq1) - 1.96 * coef_table[, "Std. Error"])
upper <- exp(coef(model_rq1) + 1.96 * coef_table[, "Std. Error"])

# Create data frame for forest plot
forest_data <- data.frame(
  variable = names(OR),
  OR = OR,
  lower = lower,
  upper = upper
)

# Remove intercept since it is not interpreted as a risk factor
forest_data <- forest_data[forest_data$variable != "(Intercept)", ]

# Rename terms for plotting
forest_data$variable <- dplyr::recode(
  forest_data$variable,
  "HadDiabetes_binaryYes" = "Diabetes: Yes",
  "SexMale" = "Sex: Male",
  "BMI" = "BMI (continuous)",
  "SmokerStatusFormer smoker" = "Smoking: Former",
  "SmokerStatusCurrent smoker - now smokes every day" =
    "Smoking: Current (daily)",
  "SmokerStatusCurrent smoker - now smokes some days" =
    "Smoking: Current (some days)"
)
# Rename age categories using gsub()
forest_data$variable <- gsub(
  "AgeCategory",
  "Age: ",
  forest_data$variable
)

library(ggplot2)

# Forest plot of adjusted odds ratios and 95% confidence intervals
ggplot(
  forest_data,
  aes(
    x = OR,
    y = reorder(variable, OR)   # Order variables by odds ratio
  )
) +
  geom_point(size = 2, color = "steelblue") +
  # Add 95% confidence interval bars
  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    height = 0.2
  ) +
  # Reference line at OR = 1 (no association)
  geom_vline(
    xintercept = 1,
    linetype = "solid",
    color = "red"
  ) +
  # Display odds ratios on a log scale for easier comparison
  # effects below and above 1 are shown symmetrically
  scale_x_log10() +
  labs(
    title = "Forest Plot of Adjusted Odds Ratios",
    subtitle = "Primary Logistic Regression Model",
    x = "Odds Ratio (95% CI, log scale)",
    y = "Variable"
  ) +
  theme_minimal()
```

## Secondary Models

For the second research question, physical activity remained
independently associated with heart disease after adjustment for age and
BMI in the multivariable logistic regression model. Respondents who
reported engaging in physical activity had lower odds of heart disease
compared with those who did not (OR = 0.60, 95% CI: 0.58–0.62, p \<
0.001). BMI showed a small but statistically significant positive
association with heart disease (OR = 1.02 per unit increase), and the
age gradient remained strong across all categories.

For the third research question, the association between smoking status
and heart disease varied across age groups in the multivariable logistic
regression model, providing evidence of effect modification by age.
Specifically, the interaction terms for current daily smoking in older
age categories were statistically significant and indicated a weaker
association between smoking and heart disease in older adults compared
with younger adults. This was most evident in ages 70–74 (OR = 0.28, 95%
CI: 0.14–0.64), 75–79 (OR = 0.21, 95% CI: 0.10–0.49), and 80 years or
older (OR = 0.18, 95% CI: 0.08–0.42). These interaction effects below 1
suggest that the increase in odds of heart disease associated with
current daily smoking is smaller in older age groups relative to the
reference age category of 18-24.

For the fourth research question, general health status remained
strongly associated with heart disease in the multivariable logistic
regression model after adjustment for age, sex, BMI, smoking status,
diabetes, and physical activity. Compared with excellent health, the
odds of heart disease increased steadily across categories (very good:
OR = 1.67; good: OR = 3.04; fair: OR = 5.66; poor: OR = 9.69; all p \<
0.001). Although these estimates were lower than in unadjusted analyses,
the association remained strong after adjustment.

## Model Diagnostics

Multicollinearity among covariates was assessed using pairwise
correlations. All pairwise correlations were below 0.45
(@fig-correlation) with the strongest observed between physical health
days and difficulty walking (r = 0.43). Diabetes showed modest positive
correlations with heart disease (r = 0.15) and BMI (r = 0.20),
consistent with the known relationship between diabetes, obesity, and
cardiovascular disease and supporting BMI's inclusion as a confounder in
the primary model. No correlations were large enough to indicate
concerning multicollinearity within the logistic regression framework.

```{r}
#| label: fig-correlation
#| fig-cap: "Correlation Matrix of Numeric and Binary Covariates"
#| echo: false

# Convert binary Yes/No variables to numeric for correlation converted to 1/0

cor_vars <- heart |>
  mutate(
    PhysicalActivities_num = if_else(PhysicalActivities == "Yes", 1, 0),
    AlcoholDrinkers_num = if_else(AlcoholDrinkers == "Yes", 1, 0),
    HadDepressiveDisorder_num = if_else(HadDepressiveDisorder == "Yes", 1, 0),
    DifficultyWalking_num = if_else(DifficultyWalking == "Yes", 1, 0),
    HadDiabetes_num = if_else(HadDiabetes_binary == "Yes", 1, 0),
    Sex_num = if_else(Sex == "Male", 1, 0)
  ) |>
  # Keep variables used in the correlation analysis
select(
  HeartDisease, BMI, SleepHours, PhysicalHealthDays, MentalHealthDays,
  # binary variables converted to 0/1
  PhysicalActivities_num, AlcoholDrinkers_num, HadDepressiveDisorder_num,
  DifficultyWalking_num, HadDiabetes_num, Sex_num
)

# Compute the correlation matrix
cor_matrix <- cor(cor_vars, use = "complete.obs")

# Plot the correlation heatmap
corrplot(
  cor_matrix,
  method = "color",
  # Show as an upper triagular matrix
  type = "upper",

  # Smaller labels for variables
  tl.cex = 0.6,
  tl.col = "black",

  # Smaller correlation numbers
  number.cex = 0.5,
  addCoef.col = "black",
  
  # Rotate labels for more space
  tl.srt = 45,

  col = colorRampPalette(c("royalblue3", "white", "red3"))(200),
  title = "Correlation Matrix of Numeric and Binary Variables",
  # Increase margins slightly
  mar = c(0, 0, 2, 0)
)
```

The Hosmer-Lemeshow goodness-of-fit test was statistically significant
($\chi^2$ = 99.41, df = 8, p \< 0.001), suggesting the model doesn't fit
the data perfectly. This indicates some lack of fit between observed and
predicted probabilities. Given the large sample size (n = 244,039), this
result was expected, as even small differences between observed and
predicted values can produce statistically significant results in large
datasets. Deviance and Pearson residual plots (@fig-residuals) showed
the two-band structure typical of logistic regression with a binary,
imbalanced outcome, with residual magnitude decreasing as fitted
probabilities increased, indicating relatively better fit among
observations assigned higher predicted probabilities of heart disease.

```{r}
#| label: fig-residuals
#| fig-cap: "Deviance and Pearson Residuals vs. Fitted Values"
#| fig-subcap: 
#|   - "Deviance Residuals"
#|   - "Pearson Residuals"
#| layout-ncol: 2
#| echo: false

ggplot(model_rq1_diag, aes(x = fitted, y = deviance_resid)) +
  geom_point(alpha= 0.1, size = 1) +
  geom_hline( yintercept = 0, linetype = "dashed", color = "royalblue") +
  labs(
    title = "Deviance residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Deviance Residuals"
  ) +
  theme_minimal()

ggplot(model_rq1_diag, aes(x = fitted, y = pearson_resid)) +
  geom_point(alpha= 0.1, size = 1) +
  geom_hline( yintercept = 0, linetype = "dashed", color = "purple") +
  labs(
    title = "Pearson residuals vs Fitted Values",
    x = "Fitted Values",
    y = "Pearson Residuals"
  ) +
  theme_minimal()
```

Influence diagnostics identified 13,471 observations (5.52%) exceeding
the standard Cook's distance cutoff of 4/n at around
$1.64 \times 10^{-5}$ (@fig-cooks). Refitting the model after excluding
these observations resulted in coefficients shrinking toward the null
(OR = 1) and odds ratios moving toward 1, indicating that these
observations contributed meaningful information to the model rather than
disproportionately distorting it. All observations were therefore
retained in the final model. Leverage diagnostics identified 21,713
observations (8.90%) exceeding the standard hat value cutoff of 2p/n at
0.0001557128 (@fig-hatvalues). However, leverage alone does not indicate
whether an observation meaningfully affects model estimates, and this
role is better captured by Cook's distance.

Taken together, these diagnostic assessments provided no evidence of
substantial violations that would compromise interpretation of the
primary logistic regression model.

```{r}
#| label: fig-cooks
#| fig-cap: "Cook's Distance for Primary Model"
#| echo: false

# Cutoff value 
n <- nrow(heart_complete)
cutoff <- 4 / n

# Cook's Distance Plot
ggplot(
  influence_measures,
  aes(x = seq_along(cooks), y = cooks)
) +
  geom_point(alpha = 0.2, size = 0.3) +
  geom_hline(
    yintercept = cutoff,
    linetype = "solid",
    color = "red"
  ) +
  labs(
    title = "Cook's Distance for Primary Model",
    x = "Observation Index",
    y = "Cook's Distance"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-hatvalues
#| fig-cap: "Hat Values for Primary Model"
#| echo: false

# Hat cutoff value
hat_cutoff <- 2 * length(coef(model_rq1)) / n

# Hat value plot
ggplot(influence_measures, aes(x = seq_along(hat), y = hat)) +
  geom_point(alpha = 0.2, size = 0.3) +
  geom_hline(
    yintercept = hat_cutoff,
    linetype = "solid",
    color = "red"
  ) +
  labs(
    title = "Hat Values for Primary Model",
    x = "Observation Index",
    y = "Hat Value"
  ) +
  theme_minimal()
```

## Model Performance

Model discrimination was evaluated using receiver operating
characteristic (ROC) curve analysis. The primary model demonstrated good
discrimination between respondents with and without heart disease (AUC =
0.786), while the full model, which included additional behavioral and
clinical covariates, showed improved discrimination (AUC = 0.823)
(@fig-roc). Although the full model achieved higher discrimination, the
primary model remained the focus of this analysis, as it was specified
to directly address the primary research question regarding diabetes and
heart disease after adjustment for age, sex, BMI, and smoking status.

```{r}
#| label: fig-roc
#| fig-cap: "ROC Curves for Primary and Full Models"
#| echo: false

ggroc(list("Primary Model" = roc_rq1, "Full Model" = roc_full)) +
  geom_abline( intercept = 1, slope = 1,
               linetype = "dashed", color = "purple4") +
  # Reference line representing random classification of heart disease status
  labs(title = "ROC Curves: Primary vs. Full Model",
       x = "Specificity",
       y= "Sensitivity") +
  theme_minimal()
```

Given the substantial class imbalance in the outcome variable, the
default classification threshold (0 = No heart disease, 1 = Yes heart
disease) of 0.50 was evaluated alongside an optimal threshold identified
using Youden’s index. At the default threshold, the model correctly
identified only 1 of 13,379 heart disease cases (sensitivity = 0.0001),
while correctly classifying nearly all non-cases (specificity = 0.999),
confirming that the default threshold is inappropriate for this outcome
prevalence (@tbl-cutoff-default). At the Youden-optimal threshold
(0.056), sensitivity improved substantially to 0.772, correctly
identifying 10,334 of 13,379 heart disease cases, with a corresponding
decrease in specificity to 0.667 (@tbl-cutoff-opt). This threshold
provides a better balance between sensitivity and specificity given the
class imbalance in the data.

```{r}
#| label: tbl-cutoff-default
#| tbl-cap: "Confusion Matrix Using Default Cutoff (0.50); 1 = Yes (Heart Disease), 0 = No"
#| echo: false

# Predicted probabilities
pred_prob <- predict(model_rq1, type = "response")

# Confusion matrix using default cutoff of 0.50
pred_50 <- ifelse(pred_prob >= 0.50, 1, 0)

confusion_50 <- table(
  Predicted = pred_50,
  Actual = heart_complete$HeartDisease
)

# Create formatted table 
knitr::kable(
  as.data.frame.matrix(confusion_50)
)
```

```{r}
#| label: tbl-cutoff-opt
#| tbl-cap: "Confusion Matrix Using Optimal Cutoff (0.056); 1 = Yes (Heart Disease), 0 = No"

#| echo: false

# Extract optimal cutoff
opt_cutoff <- as.numeric(youden["threshold"])

# Confusion matrix using optimal cutoff
pred_opt <- ifelse(pred_prob >= opt_cutoff, 1, 0)

confusion_opt <- table(
  Predicted = pred_opt,
  Actual = heart_complete$HeartDisease
)

knitr::kable(
  as.data.frame.matrix(confusion_opt)
)
```

Internal validation was performed using bootstrap resampling with 200
iterations (@fig-bootstrap). The apparent AUC from the original model
was around 0.786, and the bootstrap-derived mean AUC was nearly
identical, resulting in an optimism estimate of 0.00016. The small
optimism estimate suggests little evidence of overfitting, indicating
that the model would be expected to perform similarly in comparable
populations.

These findings indicate that the primary model provides reliable
discrimination with little evidence of overfitting, supporting its use
for evaluating the association between diabetes and heart disease.

```{r}
#| label: fig-bootstrap
#| fig-cap: "Bootstrap Distribution of AUC (200 Resamples)"
#| echo: false

# Apparent AUC
apparent_auc <- as.numeric(auc(roc_rq1))

# Distribution of bootstrap AUC values
boot_auc_df <- data.frame(AUC = boot_auc)

# Plot using ggplot
ggplot(boot_auc_df, aes(x = AUC)) +
  geom_histogram(
    bins = 20,
    color = "black",
    fill = "steelblue"
  ) +
  # Add a dashed reference line for the apparent AUC
  geom_vline(
    xintercept = apparent_auc,
    color = "red",
    linetype = "dashed"
  ) +
  labs(
    title = "Distribution of Bootstrap AUC Values",
    subtitle = "200 Bootstrap Samples",
    xlab = "AUC",
    ylab = "Frequency"
  ) +
  theme_minimal()
```

## Subgroup and Sensitivity Analyses

To assess whether the association between diabetes and heart disease
differed by sex, the primary model was refit separately for female and
male respondents, adjusting for age, BMI, and smoking status. The
association between diabetes and heart disease was observed in both
subgroups. Among female respondents, diabetes was associated with
approximately 2.63 times the odds of heart disease (95% CI: 2.47–2.81)
compared with non-diabetic females. Among male respondents, diabetes was
associated with approximately 2.17 times the odds of heart disease (95%
CI: 2.06–2.29) compared with non-diabetic males (@tbl-subgroup). The
point estimate was somewhat higher among females, suggesting a possibly
stronger association between diabetes and heart disease in this
subgroup.

```{r}
#| label: tbl-subgroup
#| echo: false

# Female diabetes OR
female_coef <- summary(model_female)$coefficients

female_or <- exp(coef(model_female)["HadDiabetes_binaryYes"]) 

female_lower <- exp(coef(model_female)["HadDiabetes_binaryYes"] - 
                   1.96 * female_coef["HadDiabetes_binaryYes", "Std. Error"])

female_upper <- exp(coef(model_female)["HadDiabetes_binaryYes"] + 
                   1.96 * female_coef["HadDiabetes_binaryYes", "Std. Error"])

# Male diabetes OR
male_coef <- summary(model_male)$coefficients

male_or <- exp(coef(model_male)["HadDiabetes_binaryYes"]) 

male_lower <- exp(coef(model_male)["HadDiabetes_binaryYes"] - 
                   1.96 * male_coef["HadDiabetes_binaryYes", "Std. Error"])

male_upper <- exp(coef(model_male)["HadDiabetes_binaryYes"] + 
                   1.96 * male_coef["HadDiabetes_binaryYes", "Std. Error"])

# Create comparison table for ORs and CIs for both sexes
subgroup_results <-data.frame( Sex = c("Female", "Male"),
                               OR = c(female_or, male_or),
                               CI_Lower = c(female_lower, male_lower),
                               CI_Upper = c(female_upper, male_upper)
                               )
# modify columns with mutate()
subgroup_results |>
  mutate(OR = round(OR, 3),
         CI_Lower = round(CI_Lower, 3),
         CI_Upper = round(CI_Upper, 3),
         ) |>
  # Display subgroup results in a formatted table
  knitr::kable(
    caption = "Association Between Diabetes and Heart disease Stratified by Sex")
```

To evaluate the impact of excluding gestational diabetes cases from the
primary analysis, a sensitivity analysis was conducted using the full
sample with gestational diabetes coded as a separate category. The
adjusted odds ratio for diabetes was nearly identical between the
primary model (OR = 2.339, 95% CI: 2.25–2.44) and the sensitivity model
(OR = 2.34, 95% CI: 2.25–2.44) (@tbl-sensitivity), indicating that the
exclusion of gestational diabetes cases did not meaningfully affect the
primary findings.

Together, these subgroup and sensitivity analyses support the robustness
of the primary finding that diabetes is independently associated with
higher odds of heart disease, while also suggesting that this
association may vary modestly by sex.

```{r}
#| label: tbl-sensitivity
#| echo: false

sensitivity_coeff <- summary(model_sensitivity)$coefficients

# Primary model OR and 95% CI
# With HadDiabetes_binary (No/Yes)
primary_or <- exp(coef(model_rq1)["HadDiabetes_binaryYes"]) 

primary_lower <- exp(coef(model_rq1)["HadDiabetes_binaryYes"] - 
                   1.96 * summary(model_rq1)$coefficients["HadDiabetes_binaryYes", "Std. Error"])

primary_upper <- exp(coef(model_rq1)["HadDiabetes_binaryYes"] + 
                   1.96 * summary(model_rq1)$coefficients["HadDiabetes_binaryYes", "Std. Error"])

# Sensitivity model OR and 95% CI
# With HadDiabetes_3level (No/Yes/Gestational)
sensitivity_or <- exp(coef(model_sensitivity)["HadDiabetes_3levelYes"]) 

sensitivity_lower <- exp(coef(model_sensitivity)["HadDiabetes_3levelYes"] - 
                   1.96 * sensitivity_coeff["HadDiabetes_3levelYes", "Std. Error"])

sensitivity_upper <- exp(coef(model_sensitivity)["HadDiabetes_3levelYes"] + 
                   1.96 * sensitivity_coeff["HadDiabetes_3levelYes", "Std. Error"])

# Create comparison table for ORs and CIs for Primary vs Sensitivity model
subgroup_results <- data.frame(
  Model = c(
    "Primary (gestational excluded)",
    "Sensitivity (gestational included)"
  ),
  OR_model = c(primary_or, sensitivity_or),
  CI_Lower_model = c(primary_lower, sensitivity_lower),
  CI_Upper_model = c(primary_upper, sensitivity_upper)
)

# Remove names of the variables in the formatted table
rownames(subgroup_results) <- NULL

subgroup_results |>
  mutate(
    OR_model = round(OR_model, 3),
    CI_Lower_model = round(CI_Lower_model, 3),
    CI_Upper_model = round(CI_Upper_model, 3)
  ) |>
  # Display sensitivity analysis results in a formatted table
  knitr::kable(
    caption = "Sensitivity Analysis: Diabetes OR with and without Gestational Diabetes"
  )
```

# Discussion

## Summary of Main Findings

This study examined the association between diabetes and heart disease
using 2022 BRFSS data from 244.039 U.S. adults. consistent with the
primary hypothesis, adults with diabetes had significantly higher odds
of heart disease compared to non-diabetic adults (adjusted OR = 2.34,
95% CI: 2.25–2.44) after adjusting for age, sex, BMI, and smoking
status. This association remained stable in a sensitivity analysis
including gestational diabetes cases and was observed in both female and
male subgroups, though the point estimate was somewhat higher among
females.

Findings from the secondary analyses were also consistent with their
respective hypotheses. Physical inactivity was associated with higher
odds of heart disease after adjusting for BMI and age. A significant
interaction between smoking and age was observed indicating that the
association between current daily smoking and heart disease varied
across groups. The effect of current daily smoking on the odds of heart
disease was greater in younger adults than in older adults. General
health status showed a clear ordinal gradient, with progressively higher
odds of heart disease as self-reported health worsened from very good to
poor, relative to excellent health, even after adjustment for behavioral
and clinical covariates. The primary model demonstrated good
discrimination (AUC = 0.786) with minimal evidence of overfitting based
on bootstrap internal validation.

## Comparison to Existing Literature

The adjusted odds ratio for diabetes in the primary model was 2.34 (95%
CI: 2.25–2.43), meaning that respondents with diabetes had more than
twice the odds of heart disease compared with those without diabetes
after adjustment for age, sex, BMI, and smoking status. This result is
similar to the findings of Yang et al., who reported an adjusted odds
ratio of 2.09 (95% CI: 1.72–2.54) using 2015 BRFSS data [@yang2019].
Their estimate was slightly lower, which may be due to differences in
the variables included in the model. In addition to demographic factors,
Yang et al. adjusted for factors such as stroke, depression, diet, and
physical activity. Despite these differences, both studies found a
strong positive association between diabetes and heart disease.

For physical Activity and heart disease, Shah et al. combined smoking,
alcohol use, and physical inactivity into a single behavioral risk score
and found that adults with two or more risk factors had 1.53 times the
odds of heart attack compared with those with no risk factors
[@shah2025]. Although the present study examined physical activity
separately, the results point in the same direction. Physically active
adults had lower odds of heart disease (OR = 0.60), indicating a
protective effect of physical activity. Equivalently, physical
inactivity remained positively associated with heart disease risk,
consistent with the findings of Shah et al.

For smoking and age, Yathindra et al. (2025) found that younger smokers
aged 18 to 24 had the highest odds of angina (OR = 4.50), with the
association decreasing steadily across age groups to an OR of 1.11 among
adults aged 65 and older [@yathindra2025]. A similar pattern was
observed in this research question, where the association between
smoking and heart disease became weaker in older age groups. One
difference is that Yathindra et al. used angina as the outcome rather
than heart disease, which may help explain their lower overall odds
ratios. Despite this difference, both studies suggest that smoking has a
stronger relative association with cardiovascular disease among younger
adults than among older adults.

None of the studies reviewed specifically examined self-reported general
health status as a predictor of heart disease using BRFSS data, which
makes direct comparisons difficult. In the present study, the odds of
heart disease increased steadily as self-reported health worsened,
ranging from 1.67 times higher odds among respondents reporting very
good health to 9.69 times higher odds among those reporting poor health,
compared with respondents reporting excellent health. This suggests that
self-rated health may be a useful indicator of cardiovascular risk.
Additional studies would be needed to determine whether a similar
pattern is observed in other BRFSS years or national health datasets.

The age gradient observed here is consistent with published BRFSS
literature. Shah et al. reported heart attack ORs increasing from 1.91
at age 25–34 to 11.22 at age 65 and older, both lower than the estimates
here, which likely reflects the broader age groupings they used
[@shah2025]. The adjusted OR for male sex in this study (OR = 2.00 in
primary model) is close to the 2.15 reported by Shah et al . and
consistent with Leal et al., who found sex to be a significant predictor
of CHD in 2020 BRFSS data [@shah2025; @leal2024].

Comparisons across studies should be interpreted with caution because
the analyses were conducted using different BRFSS survey years (2015,
2020, 2022, 2023, and 2025), different cardiovascular outcome
definitions (heart attack, angina, or coronary heart disease), and
different sets of adjustment variables. Despite these differences, the
overall findings were consistent, with diabetes, smoking, physical
inactivity, older age, and male sex all associated with higher odds of
cardiovascular disease among U.S. adults.

## Limitations

Limitations should be considered when interpreting the findings of this
study.

Although the multivariable model adjusted for age, sex, BMI, smoking
status, and other covariates, unmeasured confounders may still influence
the observed associations. Variables such as cholesterol levels, blood
pressure, family history of heart disease, and medication use were not
available in this dataset and could account for part of the association
between diabetes and heart disease. As a result, the adjusted odds
ratios reported here should not be interpreted as causal estimates,
since unmeasured factors may contribute to the observed association
between diabetes and heart disease.

All exposures and the outcome of heart disease were based on
self-reported survey responses. Respondents were asked whether they had
ever been told by a doctor that they had a heart attack, had diabetes,
or engaged in physical activity, among other questions. Self-reported
data are subject to recall bias and social desirability bias, and may
result in misclassification of both the exposure and outcome.
Individuals with undiagnosed diabetes or heart disease would be
classified as not having these conditions, which could weaken the
observed relationship between diabetes and heart disease and lead to an
underestimate of the true association.

The BRFSS is a telephone-based survey, which may underrepresent
populations without reliable phone access, including lower-income
adults, those experiencing homelessness, and non-English speakers. If
these groups differ systematically in their cardiovascular risk
profiles, the study sample may not fully representative of the U.S.
adult population, and findings may not generalize to all subgroups
equally.

The BRFSS also collects information at a single point in time, so it is
not possible to determine whether diabetes occurred before heart
disease. Because the timing of these conditions is unknown, the observed
association cannot be interpreted as a cause-and-effect relationship.
For example, some individuals may have been diagnosed with diabetes only
after receiving medical care for heart disease. Longitudinal studies
that follow individuals over time would be needed to determine whether
diabetes increases the risk of developing heart disease.

The primary exposure, diabetes status, was recoded into a binary
variable by excluding 1,983 respondents who reported gestational
diabetes and combining pre-diabetes with no diabetes. Although the
sensitivity analysis showed that this decision did not meaningfully
change the primary odds ratio (2.339 vs. 2.340), diabetes status was
simplified into broad categories that may not fully reflect differences
in disease severity or clinical history.

Taken together, these limitations suggest that even though the link
between diabetes and heart disease is consistent across multiple
sensitivity checks, it should still be interpreted as an association
rather than proof of causation. It should be considered together with
other cross-sectional and longitudinal studies.

## Public Health Implications

Diabetes was strongly associated with heart disease in this study, with
adults with diabetes having 2.34 times the odds of heart disease
compared with those without diabetes after adjustment for age, sex, BMI,
and smoking status. Given the high prevalence of diabetes in the United
States, this finding highlights the importance of diabetes prevention,
early detection, and effective disease management as part of efforts to
reduce the burden of cardiovascular disease.

Several modifiable behavioral factors were also associated with heart
disease. Current smoking was linked to higher odds of heart disease, and
physically active adults had lower odds of heart disease than inactive
adults. These findings support existing public health initiatives that
encourage smoking prevention and regular physical activity as strategies
for improving cardiovascular health and reducing disease risk.

The strong gradient observed between self-reported general health and
heart disease suggests that a simple self-rated health measure may
provide useful information about cardiovascular risk. Individuals
reporting fair or poor health had substantially higher odds of heart
disease compared to those reporting excellent health. Because self-rated
health is easy and inexpensive to collect, it may be a useful tool for
identifying populations that could benefit from additional
cardiovascular risk assessment and prevention efforts.

These findings support continued efforts to improve diabetes prevention,
early detection, and disease management. Public health strategies that
promote healthy lifestyles and cardiovascular risk reduction among
adults with diabetes may help reduce the burden of heart disease and
improve population health outcomes.

## Conclusion

This study examined the association between diabetes and heart disease
using 2022 BRFSS data from 244,039 U.S. adults, finding that diabetes
was independently associated with more than twice the odds of heart
disease after adjustment for age, sex, BMI, and smoking status. This
finding was robust across a sensitivity analysis including gestational
diabetes cases and across sex-stratified subgroup analyses, and was
consistent with prior BRFSS-based research. Secondary analyses further
showed that physical inactivity, current smoking, and poorer
self-reported general health were each independently associated with
higher odds of heart disease, while the association between smoking and
heart disease weakened with increasing age.

These findings reinforce diabetes as a key modifiable risk factor for
cardiovascular disease and support continued public health emphasis on
diabetes prevention, early detection, and management as strategies for
reducing cardiovascular risk at the population level. While the
cross-sectional design being at one point in time and reliance on
self-reported data limit causal interpretation, the consistency of these
findings with existing literature strengthens confidence in the observed
associations. Future research using longitudinal data, which follows the
same individuals over time and measures their health status at multiple
time points, and more detailed clinical measures would help clarify how
diabetes develops in relation to heart disease and further inform
targeted prevention strategies.

# References

::: {#refs}
:::
